Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International Students

### 2024-02-10 demystify.tex
updates

The most effective evidence of research ability is having \textbf{published papers in reputable international journals or conferences}.
Having papers published is a sign that the applicant has successfully involved in research.

@@

However, it is understandable that many students do not have the opportunities to publish in top places. Thus, general writings, even those under submissions or rejected, would still be good (and much better than nothing).  But be sure to upload your papers with your application and talk about them in your statement (see \autoref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still mention them in your statement and explain why they are good.

@@

Most PhD programs will require at least \textbf{two LORs}. Having a letter from an internationally recognized researcher can greatly strengthen your application. However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS).

@@

\paragraph{Waiving your Right}  Choosing not to look at a reference letter is pretty standard in school and job applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writers, and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

@@

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning about the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining acceptance to top PhD programs in the US. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), these reference letters are generally helpful as they have helped many students getting into top PhD programs in the US.

@@

\section{Interview}\label{sec:interview}

Many PhD programs have the tradition of interviewing applicants. The most common case is that a professor is interested in working with a student and wants to chat with them to make decision (e.g., to offer them RA).
The professor might include one of their current students as part of the interview.
In some case, the interview is done by several professors, e.g., to see if a student would fit in their group or to recruit a very strong student to their program.

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. A professor might want to know more about your research background or even ask you to read a paper (so be sure to do it!). Finally, during the interview, you will have the opportunity to ask questions about the university. It's essential to prepare some thoughtful questions, similar to a job interview.

\begin{commentbox}[Vu:]
  At GMU, we are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potentials and want to advocate for their admission. Without the interview, such applications may be more likely to be rejected.
\end{commentbox}

In short, getting an interview is a good sign; it means that someone is considering you. If we are not interested in your application, we will not proceed with an interview.
However, do not worry much if you do not get an interview! For example, at GMU, most of our admitted students with TA do not go through any interviews.

@@

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. In fact, if a faculty is not in CS, it is unlikely that they can be the \emph{main} adviser of a CS PhD student. They may \emph{co-advise} or be in the PhD dissertation committee, but your main adviser will need to be a tenured or tenure-track faculty in CS. If in doubt, you should check with the CS department for their requirements.  For this specific reason,  CSRankings includes only tenure line faculty who can advise CS PhD students. I also have compiled a \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{list} of Vietnamese faculty who can advise PhD students. \autoref{sec:faculty-types} talks more about who can serve as your PhD adviser.

@@

\autoref{tab:cost} shows the budget breakdown for a GRA per year (this level of details is what faculty actually uses when applying for grants).
These numbers are based on my experience at public universities in the US.  Private universities may have different numbers.  For simplicity, in this table I will assume the department has a 9-month stipend of \$27000 (GMU actually pays more) and therefore a 3-month summer of \$9000. I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., MIT charges around \$50K) and a 58.9\% rate on \emph{indirect cost}, which is what GMU charges for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

@@

This section lists some cultural issues that international students might want to pay attention to. These issues might not specific to just the PhD admission process but in general.

@@

\paragraph{Accepting an offer} If you decide to accept an offer, you can say something like: \emph{``Thank you for the offer.  I am very excited about it and would like to accept it.  I look forward to working with you.  Can you please send me more details about the offer and what to do next?''} .  Here you are not only accepting the offer but also asking for more details.  This is perfectly fine and professors will understand and appreciate your enthusiasm.  They will likely send you more details about the offer and what to do next.  If you decide to accept an offer, you should do so quickly.

@@

\paragraph{Postponing an offer} If you need more time to decide, you can ask for more time.  You can say something like: \emph{``Thank you for the offer.  I am very excited about it.  However, I am still waiting for other offers and need more time to decide.  Would it be possible to postpone the decision for a few weeks?''}.  This is perfectly fine and professors will understand and might even appreciate your honesty.  They will likely give you a few weeks to decide.  If you need more time, you can ask for more time.  But do not ask for too much time, e.g., more than a month.  You also should not postpone the offers multiple times, which will likely annoy people.

\paragraph{Declining an offer} If you decide to decline or reject an offer, you can say something like: \emph{``Thank you for the offer.  I am very excited about it.  However, I have decided to accept another offer.  I appreciate your time and consideration.  I hope we can work together in the future.''}  Professors will understand and will likely wish you good luck.  If you decide to reject an offer, you should do so quickly.

>>>>

The most concrete evidence of research ability is having \textbf{published papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has successfully involved in research.

@@

However, it is understandable that many students do not have the opportunities to publish in top places. Thus, general writings, even those under submissions or rejected, would still be good (and much better than nothing).  But be sure to upload your papers with your application and talk about them in your statement (see \autoref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still upload them, mention them in your statement, and explain why they are good.

@@

Most PhD programs will require at least \textbf{two LORs}. Having a letter from an internationally recognized researcher can \textbf{greatly} strengthen your application. In fact, it is arguable at many top CS programs, strong LoRs from well-known researchers are even more important than published papers.

However, obtaining such letters
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS).

@@

\paragraph{Waiving Your Right}  Choosing not to look at a reference letter is pretty standard in school and job applications. When you waive your right to see the letter, it adds a layer of trust, showing you're confident in your choice of referees and that you're not trying to twist their words. It's also about keeping things open and honest between you and your letter writers, and encourages them to be real about your strengths and qualifications. Plus, it keeps things private.

@@

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning about the \href{https://vef2.org/}{VEF2.0 program}, which has helped many good students in gaining admission to top CS PhD programs. VEF2.0 follows an interesting model where US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), these reference letters are generally effective as they have helped many students getting into top PhD programs in the US.

@@

\section{Interview and the Waiting Game}\label{sec:interview}

After you submitted your applications, the waiting game begins! For many students, this is an excruciating time. This section provides some information and tips to help you get through this time.

% but you should try to relax and focus on your study and work.  You can also use this time to improve your profile, e.g., by working on research projects, publishing papers, or improving your connections for better LoR writers.

\paragraph{Interviews} Many PhD programs have the tradition of interviewing applicants (just like job applications). The most common case is that a professor is interested in working with a student and want to chat with them to make decision (e.g., to offer them RA).
The professor might include one of their current students as part of the interview.

In some case, the interview is done by several professors, e.g., to see if a student fits to their group or to recruit a very strong student to their program. The interview can also be done by Adcom members to see if the student fit their program.

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. A professor might want to know more about your research background or even ask you to read a paper (so be sure to do it!). During the interview, you will have the opportunity to ask questions about the university. It's essential to prepare some thoughtful questions, similar to a job interview.

\begin{commentbox}[Vu:]
  At GMU, faculty are encouraged to interview candidates. For very strong candidates, the interview is actually to recruit them.  In some cases a faculty interviews a candidate that they see potentials and want to advocate for their admission. Without the interview, such applications may be more likely to be rejected.
  \tcblower
  In short, getting an interview is a good sign; it means that someone is considering you. If we are not interested in your application, we will not proceed with an interview.

\paragraph{Not getting interviews} While it is generally good to get interview, do not feel so despair if you do not get any.  Many programs do not have the tradition of interviewing applicants. For example, at GMU, most of our admitted students with TA do not go through any interviews.

\paragraph{Timeline} The timeline for interviews and notification letters varies.  Faculty setup interviews based on their (very busy) schedule. So do not be surprised if you get an interview invitation in the last minute (also don't get surprised if your interview is on a Uber ride!). And again, some schools do not do interviews at all.

Some schools send out admission letters in batches, some do rolling admission, and some do not send anything out (e.g., you're on their wait list).  However, in general, you should hear back from most schools by around March.

In short, you should not worry too much about the timeline. Also do not send emails asking about interviews or status.

\paragraph{Updating your profile} In general, you should not send emails to update your profile.  However, if you have new publications or other achievements, you can send an email asking to update your profile.

@@

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. In fact, if a faculty is not in CS, it is unlikely that they can be the \emph{main} adviser of a CS PhD student. They may \emph{co-advise} or be in the PhD dissertation committee, but your main adviser will need to be a tenured or tenure-track faculty in CS. If in doubt, check with the CS department for their requirements.  For this specific reason,  CSRankings includes only tenure line faculty who can advise CS PhD students. I also have compiled a \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{list} of Vietnamese faculty who can advise PhD students. \autoref{sec:faculty-types} talks more about who can serve as your PhD adviser.

@@

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US.  Private universities may have different numbers.  For simplicity, in this table I will assume the department has a 9-month stipend of \$27000 (GMU actually pays more) and therefore a 3-month summer of \$9000. I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., MIT charges around \$50K) and a 58.9\% rate on \emph{indirect cost}, which is what GMU charges for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

@@

This section lists some cultural issues that international students might want to pay attention to. These issues apply in general and not just PhD admission.

@@

\paragraph{Accepting an offer} If you decide to accept an offer, you can say: \emph{``Thank you for the offer.  I would like to accept it and look forward to working with you.  Could you please send me more details about the offer and what to do next?''}.  Here you are not only accepting the offer but also asking for more details. The prof. will likely send you more details about the offer and what to do next.  If you decide to accept an offer, do so quickly.

@@

\paragraph{Postponing an offer} If you need more time to decide, you can ask for more time: \emph{``Thank you for the offer.  I am very excited about it.  However, I am still waiting for other offers and need more time to decide.  Would it be possible to postpone the decision for a few weeks?''}.  This is perfectly fine and professors will understand and might even appreciate your honesty.  They will likely give you a few weeks to decide.  If you need more time, you can ask for more time.  But do not ask for too much time, e.g., more than a month.  You also should not postpone the offers multiple times, which will annoy people.

\paragraph{Declining an offer} If you decide to decline or reject an offer, you can say: \emph{``Thank you for the offer. However, I have decided to accept another offer.  I appreciate your time and consideration.  I hope we can work together in the future.''}  Professors will understand and wish you good luck.  If you decide to reject an offer, do so quickly.
